---
title: "Sample Size Calculation"
author: "Sebastian Quirarte"
date: "September 2023"
output:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\ 

The goal of this document is to provide a template in R that can calculate the sample size needed for clinical trials and other experiments based on a range of specified values and a set type l error ($\alpha$), type two error ($\beta$), and statistical power ($1 - \beta$). A range of the means or proportions of the treatment group of interest can be established and all sample sizes calculated, stored in a data frame, and plotted. 

Group A is considered the 'new treatment group' and the sample size calculated is for EACH of the two treatment arms/groups. In other words, **total sample size is equal to double the calculated value.** 
\ 

\ 

_**Programming Language**_: R | _**Packages**_: ggplot2, ggthemes
\ 

*Source and inspiration:* HyLown Consulting LLC (2022), https://powerandsamplesize.com/

\newpage

First, let's remove previous variables as part of good practices in R, and load the necessary packages for the ggplot() and themes() functions.
```{r results = 'hide', warning=FALSE, message=FALSE}
rm(list = ls()) 
library("ggplot2") # ggplot
library("ggthemes") # theme_minimal() 
```

# Compare Two Means 

## Equality 

Equality tests aim to prove if the means of two groups are different. Suppose the two groups are 'A' and 'B', and we collect a sample from both groups -- i.e. we have two samples. We perform a two-sample test to determine whether the mean in group A ($\mu_{A}$), is different from the mean in group B ($\mu{B}$).

The hypotheses are: 

\[
  H_{0}: \mu_{A} - \mu_{B} = 0 
\]
\[
  H_{1}: \mu_{A} - \mu_{B} \neq 0
\]

Let's load up our necessary variables.
```{r}
muA = 5 # group A mean (new treatment)
muB = 10 # group B mean
sd = 10 # standard deviation
kappa = 1 # sampling ratio
alpha = 0.05 # type 1 error 
beta = 0.20 # type 2 error 
```

We can set a range of treatment group means ($\mu_{A}$) that we want to plot with their respective sample sizes. 
```{r}
range_meanA = seq(1, 8, 0.1) # (start, stop, step)
```

Let's create a data frame with this range of means and an empty column for sample sizes (n).
```{r}
df <- data.frame("meanA"=range_meanA, "n"=0)
```

This for loop calculates the sample sizes for each of the means in our selected range and adds them to our data frame. 
```{r}
# Sets counter variable to iterate through each row 
count = 1
# Calculates sample size for each mean of group A and adds value to data frame
for (i in range_meanA){
  muA = i 
  nB = ceiling((1+1/kappa)*(sd*(qnorm(1-alpha/2)+qnorm(1-beta))/(muA-muB))^2)
  df[count, 2] = nB
  count = count + 1
}
```

Here's a look at our data frame of calcualted sample sizes per treatment arm:
```{r}
head(df, 3) # shows first 3 rows
tail(df, 3) # shows last 3 rows 
```

Finally, we can create a plot of the sample sizes based on the range of means of interest.
```{r}
ggplot(df, aes(x = meanA, y = n)) + # data and x, y values
  geom_point(color="dodgerblue4") + # adds plot point color 
  theme_minimal() + ggtitle("Compare Two Means: Equality") + # theme and title
  theme(plot.title = element_text(hjust = 0.5)) + # center title 
  xlab("Group A Mean") + ylab("n per treatment") # x and y axis labels 
```

## Non-Inferiority or Superiority

Non-inferiority and superiority tests aim to prove that a new treatment is either non-inferior or superior to another treatment, by a set margin known as a margin of non-inferiority ($-\delta$) or margin of superiority ($\delta$). In this setting, we wish to test whether the mean in group 'A', $\mu_{A}$, is non-inferior/superior to the mean in group 'B', $\mu_{B}$. We collect a sample from both groups, and thus will conduct a two-sample test. The idea is that statistically significant differences between the means may not be of interest unless the difference is greater than a threshold, $\delta$. This is particularly popular in clinical studies, where the margin is chosen based on clinical judgement and subject-domain knowledge. 

The hypotheses to test are:

\[
  H_{0}: \mu_{A} - \mu_{B} \le \delta 
\]
\[
  H_{1}: \mu_{A} - \mu_{B} > \delta
\]

Let's load up our necessary variables.
```{r}
muA = 5 # group A mean (new treatment)
muB = 5 # group B mean
sd = 10 # standard deviation
delta = 5 # non-inferiority (-) or superiority (+) margin
kappa = 1 # sampling ratio
alpha = 0.05 # type 1 error 
beta = 0.20 # type 2 error 
```

We can set a range of treatment group means ($\mu_{A}$) that we want to plot with their respective sample sizes. 
```{r}
range_meanA = seq(5, 8, 0.1) # (start, stop, step)
```

Let's create a data frame with this range of means and an empty column for sample sizes (n).
```{r}
df <- data.frame("meanA"=range_meanA, "n"=0)
```

This for loop calculates the sample sizes for each of the means in our selected range and adds them to our data frame. 
```{r}
# Sets counter variable to iterate through each row 
count = 1
# Calculates sample size for each mean of group A and adds value to data frame
for (i in range_meanA){
  muA = i 
  nB = ceiling((1+1/kappa)*(sd*(qnorm(1-alpha)+qnorm(1-beta))/(muA-muB-delta))^2)
  df[count, 2] = nB
  count = count + 1
}
```

\newpage

Here's a look at our data frame of calculated sample sizes per treatment arm:
```{r}
head(df, 3) # shows first 3 rows
tail(df, 3) # shows last 3 rows 
```

Finally, we can create a plot of the sample sizes based on the range of means of interest.
```{r}
ggplot(df, aes(x = meanA, y = n)) + # data and x,y values 
  geom_point(color="dodgerblue4") + # plot point color an 
  theme_minimal() + ggtitle("Compare Two Means: Superiority") + # theme & title
  theme(plot.title = element_text(hjust = 0.5)) + # center title
  xlab("Group A Mean") + ylab("n per treatment") # axis labels 
```

\newpage

If we change the sign of our margin from $\delta$ to -$\delta$, we can see this same result if applied to a non-inferiority trial instead of superiority. 
```{r}
muA = 5 # group A mean (new treatment)
muB = 5 # group B mean
sd = 10 # standard deviation
delta = -5 # non-inferiority (-) or superiority (+) margin
kappa = 1 # sampling ratio
alpha = 0.05 # type 1 error 
beta = 0.20 # type 2 error 
```

```{r}
range_meanA = seq(4, 8, 0.1) # (start, stop, step)
```

```{r}
df <- data.frame("meanA"=range_meanA, "n"=0)
```

```{r}
# Sets counter variable to iterate through each row 
count = 1
# Calculates sample size for each mean of group A and adds value to data frame
for (i in range_meanA){
  muA = i 
  nB = ceiling((1+1/kappa)*(sd*(qnorm(1-alpha)+qnorm(1-beta))/(muA-muB-delta))^2)
  df[count, 2] = nB
  count = count + 1
}
```

```{r}
head(df) #shows first 5 values
tail(df) # shows last 5 values 
```

```{r}
ggplot(df, aes(x = meanA, y = n)) + # data and x,y values 
  geom_point(color="dodgerblue4") + theme_minimal() + # plot point color + theme
  ggtitle("Compare Two Means: Non-Inferiority") + # title
  theme(plot.title = element_text(hjust = 0.5)) + # centers title 
  xlab("Group A Mean") + ylab("n per treatment") # axis labels 
```

## Equivalence 

An equivalence test is useful when we wish to test whether the means of two groups are equivalent, without concern for which group's mean is larger. Suppose we collect a sample from a group 'A' and a group 'B'; that is we collect two samples and will conduct a two-sample test. For example, we may wish to test whether a new product is equivalent to an existing, industry standard product. Here, the 'burden of proof', so to speak, falls on the new product; that is, equivalence is actually represented by the alternative, rather than the null hypothesis.

The hypotheses to test are:

\[
  H_{0}: |\mu_{A} - \mu_{B}| \ge \delta 
\]
\[
  H_{1}: |\mu_{A} - \mu_{B}| < \delta
\]

Let's load up our necessary variables 
```{r}
muA = 5 # group A mean (new treatment)
muB = 4 # group B mean
sd = 10 # standard deviation
delta = 5 # equivalence margin
kappa = 1 # sampling ratio
alpha = 0.05 # type 1 error 
beta = 0.20 # type 2 error 
```

\newpage

We can set a range of treatment group means ($\mu_{A}$) that we want to plot with their respective sample sizes. 
```{r}
range_meanA = seq(4, 7, 0.1) # (start, stop, step)
```

Let's create a data frame with this range of means and an empty column for sample sizes (n).
```{r}
df <- data.frame("meanA"=range_meanA, "n"=0)
```

This for loop calculates the sample sizes for each of the means in our selected range and adds them to our data frame. 
```{r}
# Sets counter variable to iterate through each row 
count = 1
# Calculates sample size for each mean of group A and adds value to data frame
for (i in range_meanA){
  muA = i 
  nB = ceiling((1+1/kappa)*(sd*(qnorm(1-alpha)+qnorm(1-beta/2))/(abs(muA-muB)-delta))^2)
  df[count, 2] = nB
  count = count + 1
}
```

Here's a look at our data frame:
```{r}
head(df, 3) # shows first 3 rows
tail(df, 3) # shows last 3 rows 
```

Finally, we can create a plot of the sample sizes based on the range of means of interest.
```{r}
ggplot(df, aes(x = meanA, y = n)) + # data and x,y values 
  geom_point(color="dodgerblue4") + # plot point color
  theme_minimal() + # plot theme 
  ggtitle("Compare Two Means: Equivalence") + # plot title 
  theme(plot.title = element_text(hjust = 0.5)) + # centers title 
  xlab("Mean Group A") + ylab("n per treatment") # axis labels 
```

# Compare Two Proportions

## Equality 

Equality tests aim to prove if the proportions of two groups are different. Suppose the two groups are 'A' and 'B', and we collect a sample from both groups -- i.e. we have two samples. We perform a two-sample test to determine whether the proportion in group A ($p_{A}$) is different from the proportion in group B ($p_{B}$).

The hypotheses are: 

\[
  H_{0}: p_{A} - p_{B} = 0 
\]
\[
  H_{1}: p_{A} - p_{B} \neq 0
\]

Let's load up our necessary variables.
```{r}
pA = 0.65 # group A proportion (new treatment)
pB = 0.85 # group B proportion
kappa = 1 # sampling ratio
alpha = 0.05 # type 1 error 
beta = 0.20 # type 2 error 
```


We can set a range of treatment group proportions ($p_{A}$) that we want to plot with their respective sample sizes.
```{r}
range_propA = seq(0.65, 0.78, 0.01) # (start, stop, step)
```

Let's create a data frame with this range of proportions and an empty column for sample sizes (n).
```{r}
df <- data.frame("propA"=range_propA, "n"=0)
```

This for loop calculates the sample sizes for each of the means in our selected range and adds them to our data frame. 
```{r}
# Sets counter variable to iterate through each row 
count = 1
# Calculates sample size for each proportion of group A and adds value to data frame
for (i in range_propA){
  pA = i 
  nB = ceiling((pA*(1-pA)/kappa+pB*(1-pB))*((qnorm(1-alpha/2)+qnorm(1-beta))/(pA-pB))^2)
  df[count, 2] = nB
  count = count + 1
}
```

Here's a look at our data frame:
```{r}
head(df, 3) # shows first 3 rows
tail(df, 3) # shows last 3 rows 
```

Finally, we can create a plot of the sample sizes based on the range of means of interest.
```{r}
ggplot(df, aes(x = propA, y = n)) + # data and x,y values
  geom_point(color="dodgerblue4") + # plot point color 
  theme_minimal() + # plot theme 
  ggtitle("Compare Two Proportions: Equality") + # plot title 
  theme(plot.title = element_text(hjust = 0.5)) + # centers title
  xlab("Proportion Group A") + ylab("n per treatment") # axis labels 
```

## Non-inferiority or Superiority

Non-inferiority and superiority tests aim to prove that a new treatment is either not-inferior or superior compared to another treatment, by a set margin known as a margin of non-inferiority ($-\delta$) or margin of superiority ($\delta$). In this setting, we wish to test whether the proportion in group 'A', $p_{A}$, is non-inferior/superior to the proportion in group 'B', $p_{B}$. We collect a sample from both groups, and thus will conduct a two-sample test. The idea is that statistically significant differences between the proportions may not be of interest unless the difference is greater than a threshold, $\delta$. This is particularly popular in clinical studies, where the margin is chosen based on clinical judgement and subject-domain knowledge. 

The hypotheses to test are:

\[
  H_{0}: p_{A} - p_{B} \le \delta 
\]
\[
  H_{1}: p_{A} - p_{B} > \delta
\]

Let's load up our necessary variables.
```{r}
pA = 0.85 # group A proportion (new treatment)
pB = 0.65 # group B proportion
delta = -0.10 # non-inferiority (-) or superiority (+) margin
kappa = 1 # sampling ratio
alpha = 0.05 # type 1 error 
beta = 0.20 # type 2 error 
```

We can set a range of treatment group proportions ($p_{A}$) that we want to plot with their respective sample sizes. 
```{r}
range_propA = seq(0.65, 0.85, 0.01) # (start, stop, step)
```

Let's create a data frame with this range of means and an empty column for sample sizes (n).
```{r}
df <- data.frame("propA"=range_propA, "n"=0)
```

This for loop calculates the sample sizes for each of the means in our selected range and adds them to our data frame. 
```{r}
# Sets counter variable to iterate through each row 
count = 1
# Calculates sample size for each proportion of group A and adds value to data frame
for (i in range_propA){
  pA = i 
  nB = ceiling((pA*(1-pA)/kappa+pB*(1-pB))*((qnorm(1-alpha)+qnorm(1-beta))/(pA-pB-delta))^2)
  df[count, 2] = nB
  count = count + 1
}
```

Here's a look at our data frame:
```{r}
head(df, 3) # shows first 3 rows
tail(df, 3) # shows last 3 rows 
```

Finally, we can create a plot of the sample sizes based on the range of means of interest.
```{r}
ggplot(df, aes(x = propA, y = n)) + # data and x,y values
  geom_point(color="dodgerblue4") + # plot point color 
  theme_minimal() + # plot theme 
  ggtitle("Compare Two Proportions: Non-Inferiority") + # plot title 
  theme(plot.title = element_text(hjust = 0.5)) + # centers title
  xlab("Proportion Group A") + ylab("n per treatment") # axis labels 
```

If we change the values we can see this same result if applied to a superiority trial instead of non-inferiority.
```{r}
pA = 0.65 # group A proportion (new treatment)
pB = 0.85 # group B proportion
delta = 0.10 # non-inferiority (-) or superiority (+) margin
kappa = 1 # sampling ratio
alpha = 0.05 # type 1 error 
beta = 0.20 # type 2 error 
```

```{r}
range_propA = seq(0.65, 0.85, 0.01) # (start, stop, step)
```

```{r}
df <- data.frame("propA"=range_propA, "n"=0)
```

```{r}
# Sets counter variable to iterate through each row 
count = 1
# Calculates sample size for each proportion of group A and adds value to data frame
for (i in range_propA){
  pA = i 
  nB = ceiling((pA*(1-pA)/kappa+pB*(1-pB))*((qnorm(1-alpha)+qnorm(1-beta))/(pA-pB-delta))^2)
  df[count, 2] = nB
  count = count + 1
}
```

```{r}
head(df) # shows first 5 rows
tail(df) # shows last 5 rows 
```

```{r}
ggplot(df, aes(x = propA, y = n)) + # data and x,y values
  geom_point(color="dodgerblue4") + theme_minimal() + # plot point color + theme
  ggtitle("Compare Two Proportions: Superiority") + # plot title
  theme(plot.title = element_text(hjust = 0.5)) + # centers title
  xlab("Proportion Group A") + ylab("n per treatment") # axis labels
```

## Equivalence

An equivalence test is useful when we wish to test whether the proportions of two groups are equivalent, without concern of which group's proportion is larger. Suppose we collect a sample from a group 'A' and a group 'B'; that is we collect two samples, and will conduct a two-sample test. For example, we may wish to test whether a new product is equivalent to an existing, industry standard product. Here, the 'burden of proof', so to speak, falls on the new product; that is, equivalence is actually represented by the alternative, rather than the null hypothesis.

The hypotheses to test are:

\[
  H_{0}: |p_{A} - p_{B}| \ge \delta 
\]
\[
  H_{1}: |p_{A} - p_{B}| < \delta
\]

Let's load up our necessary variables.
```{r}
pA = 0.65 # group A proportion (new treatment)
pB = 0.85 # group B proportion
delta = 0.05 # equivalence margin
kappa = 1 # sampling ratio
alpha = 0.05 # type 1 error 
beta = 0.20 # type 2 error
```

We can set a range of treatment group proportions ($p_{A}$) that we want to plot with their respective sample sizes. 
```{r}
range_propA = seq(0.60, 0.75, 0.01) # (start, stop, step)
```

Let's create a data frame with this range of means and an empty column for sample sizes (n).
```{r}
df <- data.frame("propA"=range_propA, "n"=0)
```

This for loop calculates the sample sizes for each of the means in our selected range and adds them to our data frame. 
```{r}
# Sets counter variable to iterate through each row 
count = 1
# Calculates sample size for each proportion of group A and adds value to data frame
for (i in range_propA){
  pA = i 
  nB = ceiling((pA*(1-pA)/kappa+pB*(1-pB))*((qnorm(1-alpha)+qnorm(1-beta/2))/(abs(pA-pB)-delta))^2)
  df[count, 2] = nB
  count = count + 1
}
```

\newpage 

Here's a look at our data frame:
```{r}
head(df, 3) # shows first 3 rows
tail(df, 3) # shows last 3 rows 
```

Finally, we can create a plot of the sample sizes based on the range of means of interest.
```{r}
ggplot(df, aes(x = propA, y = n)) + # data and x,y values
  geom_point(color="dodgerblue4") + theme_minimal() + # plot point color + theme
  ggtitle("Compare Two Proportions: Equivalence") + # plot title 
  theme(plot.title = element_text(hjust = 0.5)) + # centers title
  xlab("Proportion Group A") + ylab("n per treatment") # axis labels 
```

# Version Control 

September 13, 2023 -- Version 1.1 -- spelling and added titles to plots. 
  
September 9, 2023 -- Version 1.0 -- Creation of first finished version.




